How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-20 Paper/paper.tex
some assorted changes

\subsubsection{Remixing}
Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}. Shared Scratch programs can be `remixed' by other Scratch users, which means that a copy of this program is placed in the user's own project collection, and can be then further changed. The `remix tree' of projects is public, so users can track which users remix their programs, a bit similar forking in GitHub. Contrary to forking though, changes upstream cannot be integrated back into the original project.

\begin{figure}
	\begin{center}
		\includegraphics[width=6cm]{fig/makeBlockSmall.png}
		\caption{}
		\label{fig:makeBlock}
	\end{center}
\end{figure}

\begin{figure}
	\begin{center}
		\includegraphics[width=6cm]{fig/useBlockSmall.png}
		\caption{}
		\label{fig:useBlock}
	\end{center}
\end{figure}

@@

The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs. To answer our research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch repository. The dataset is described in Section \ref{dataset}. Below, we present how we approached each research question.

@@

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrrrrrr}
		\hline
		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Sprites with code per project&5.68&1&1&2&5&525\\
		Scripts per project&17.35&1&2&5&12&3038\\
		Maximum indentation per project&4.56&1&3&3&5&488\\
		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
		LOC in Stage per project&4.80&0&0&0&3&2613\\
		LOC in Sprites per project&115.57&0&10&26&68&34613\\
		LOC in Procedures per project&34.17&0&0&0&0&20552\\
		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		McCabe CC per procedure script&3.75&1&1&2&4&183\\
		Procedures per project with procedures&11.50&1&1&2&6&847\\
		Arguments per Procedure&0.95&0&0&0&1&53\\
		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		Calls per procedure&2.14&0&1&1&2&526\\
		Scripts with calls per procedure&1.13&0&1&1&1&59\\
		Variables per project&2.06&0&0&0&1&340\\
		Scripts utilizing variable&4.97&1&1&3&5&1127\\
		Lists per project&0.55&0&0&0&0&319\\
		Conditional statements per project&10.02&0&0&0&3&5950\\
		Recursive loop statements per project&7.65&0&1&2&5&2503\\
		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms  non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

The lines of code metric was further analyzed to understand code organization. As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites. An additional 3.1\% of the total lines are found in the stage class. More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects. Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.

@@

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		\emph{RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points. Most code is written in sprites. Few projects (7.7\%) use procedures, but they use them a lot and for more complex code. There exist surprisingly large and complex projects.}
	\end{minipage}}

@@

As a first programming abstraction we investigated the use of procedures. In the dataset we found 206,799 procedures in 17,979 projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of those projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\%\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		\emph{RQ2: One third of the projects utilize variables, sometimes without initializing them. 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used. 29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist.}
	\end{minipage}}

@@

\subsubsection{Code Clones Between Sprites}
With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common. In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of procedures between scripts, only within them. So in many cases there is no other way to share the functionality than making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users needs this functionality.

@@

>>>>

\subsubsection{Procedures}
Scratch users can define their own blocks, with a pop-up screen displayed in on the left hand side of Figure \ref{fig:twoBlocks}. As shown in this Figure, Scratch users can set the name of the sprite and define input parameters. The block in Figure \ref{fig:twoBlocks} uses two parameter, one textual (string1) and one numerical (number1). A user-created block is programmed using ordinary Scratch blocks as shown on the right hand side of Figure \ref{fig:twoBlocks}.

\subsubsection{Remixing}
\todo{maybe only useful with we will talk about remixing in the remainder?}
Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}. Shared Scratch programs can be `remixed' by other Scratch users, which means that a copy of this program is placed in the user's own project collection, and can be then further changed. The `remix tree' of projects is public, so users can track which users remix their programs, a bit similar forking in GitHub. Contrary to forking though, changes upstream cannot be integrated back into the original project.

\begin{figure}
	\begin{center}
		\includegraphics[width=8cm]{fig/twoBlocks.png}
		\caption{The user interface for creating and defining user-defined blocks \todo{could be nicer}}
		\label{fig:twoBlocks}
	\end{center}
\end{figure}

@@

The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs. To answer our research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch repository. The dataset is described in Section \ref{dataset}. \todo{Below, we present how we approached each research question.}

@@

\begin{table*}[ht]
	\centering
	\begin{tabular}{lrrrrrr}
		\hline
		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
		\hline
		Sprites with code per project&5.68&1&1&2&5&525\\
		Scripts per project&17.35&1&2&5&12&3038\\
		Maximum indentation per project&4.56&1&3&3&5&488\\
		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
		LOC in Stage per project&4.80&0&0&0&3&2613\\
		LOC in Sprites per project&115.57&0&10&26&68&34613\\
		LOC in Procedures per project&34.17&0&0&0&0&20552\\
		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
		McCabe CC per procedure script&3.75&1&1&2&4&183\\
		Procedures per project with procedures&11.50&1&1&2&6&847\\
		Arguments per Procedure&0.95&0&0&0&1&53\\
		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
		Calls per procedure&2.14&0&1&1&2&526\\
		Scripts with calls per procedure&1.13&0&1&1&1&59\\
		Variables per project&2.06&0&0&0&1&340\\
		Scripts utilizing variable&4.97&1&1&3&5&1127\\
		Lists per project&0.55&0&0&0&0&319\\
		Conditional statements per project&10.02&0&0&0&3&5950\\
		Recursive loop statements per project&7.65&0&1&2&5&2503\\
		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\
		\hline
	\end{tabular}
	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
	\label{tbl-size}
\end{table*}

@@

The lines of code metric was further analyzed to understand code organization. As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC~lines of code---is written within sprites. An additional 3.1\% of the total lines are found in the stage class. More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects. Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.

@@

\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points. Most code is written in sprites. A small number of projects (7.7\%) use procedures, but they use them a lot and for more complex code. There exist surprisingly large and complex projects.
	\end{minipage}}

@@

The first method for abstraction that we investigate is the use of procedures. In the dataset we found 206,799 procedures in 17,979 projects. As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of those projects having up to 2. Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects. Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\%\% have only one (shown in Figure \ref{fig:procedurearguments}). The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.

@@

%user interativity		
\noindent
\fbox{
	\begin{minipage}{0.46\textwidth}
		RQ2: One third of the projects utilize variables, sometimes without initializing them. 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used. 29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized. Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure. Recursive procedure calls exist.
	\end{minipage}}

@@

\subsubsection{Code Clones Between Sprites}
With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common. In a sense, the Scratch users are not to blame here, as Scratch does not support the sharing of procedures between scripts, only within them. So in many cases there is no other way to share the functionality than by making a copy. We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.

@@

\subsection{Threads to validity}
A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 programs. It could be the case that programming habits of Scratch users are changing over time. However, we counterbalanced that by using a very large dataset. Furthermore, for some of the analyses we manually inspected programs \todo{Fenia, which ones?}
